The Multi-Stencil Language

### 2016-02-01 ARTICLES/IJPP/article/formalism.tex
article IJPP corrections+example Helene 01/02/2016

@@

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item An entity $e$ of a mesh $\mathcal{M}=(V,E)$ is defined as a subset of its vertices and edges, $e\subset V\cup E$.
\item A mesh entities group is denoted $E$ such that $E \in \mathcal{P}(V\cup E)$. It represents a set of entities of the same type.
\item The set of mesh entities groups of a simulation is denoted $\mathcal{E}$.
\item Finally, the mesh on which is applied a group of mesh entities $E$ is denoted $mesh(E)$
\end{itemize}
%\end{mydef}

For example, in a 2D Cartesian mesh an entity could be a cell, made of four vertices and four edges, or simply a vertex. As a result a group of cells and a group of vertices could be defined in $\mathcal{E}$. %This type can be defined as the sets containing exactly four vertices and four edges connected as a cycle. Another type of entities simply are the vertices ($V$) and can be defined as all singletons formed of a single vertex of $V$.

@@

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item A quantity (or data) is a function $\delta: E_{\delta} \mapsto V_{\delta}$ which associates each entity of a group $e\in E_{\delta}$ to a value $v\in V_{\delta}$ where $V_{\delta}$ is a data type, typically $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.
\item The set of quantities applied onto the mesh is denoted $\Delta$.
\item In the rest of this paper, the group of mesh entities on which a quantity $\delta$ is mapped is denoted $entity(\delta)=E_{\delta}$.
\end{itemize}
%\end{mydef}

Another option, closer to the applied mathematics domain, would have been to define $\delta$ as the function over $E_{\delta} \times T$; however the approach chosen for now in this work is to let the user be aware of the number of data he is using and what exactly for.

@@

In this section are considered three different types of kernel computations, stencil kernels, local kernels, and reduction kernels.

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item A computation domain $D$ is a subpart of a mesh entities group, $D \subset E \in \mathcal{E}$.
\item The set of computation domains of a numerical simulation is denoted $\mathcal{D}$.
\item A neighborhood $n$ is a function which for a given entity $e \in E_i$, returns a set of $m$ entities in $E_j$, $n : E_i \rightarrow E_j^m$. One can notice that $i = j$ is possible. Most of the time, such a neighborhood is called a stencil shape.
\item The set of neihborhood functions in a numerical simulation is denoted $\mathcal{N}$.
\end{itemize}

\begin{mydef}
A computation kernel $k$ of a numerical simulation is defined as $k(S,R,(w,D),comp)$, where 
\begin{itemize}
\item $S \in \mathcal{S}$ is the set of scalar to read, 
\item $(w,D) \in \Delta \times \mathcal{D}$ is the single data modified by the computation kernel onto a given computation domain. Thus, $w \in \Delta$ and $D \in \mathcal{D}$ is the computation domain on which $w$ is computed, $D \subset E_w=entity(w)$.
\item $R$ is the set of tuples $(r,n)$, where $r \in \Delta$ and $n \in \mathcal{N}$ is a neighborhood function such that $n : E_w \rightarrow entity(r)^m$. The neighborhood indicates which entities of $r$ will be read in order to compute a single entity of $w$. 
\item $comp$ is the numerical computation which returns a value from a set of input values, $comp: V^n \rightarrow V$.
\end{itemize}
% \item A numerical expression $\text{exp}_{w}: entity(w) \times R^n \rightarrow V_{w}$ is a function which computes the values of the written quantity $w \in \Delta$ for a given entity $e \in entity(w)=E_w$, using a set of input data $R$.
% \item A computation kernel $k$ of a numerical simulation is defined as $c(R,w,\text{exp},D)$, where $R$ is the set of data read, $w \in \Delta$ is the unique data written, $\text{exp}$ is a numerical expression, and $D \in \mathcal{D}$ is the computation domain.
% \end{itemize}
%$R$ is the set of data read, $w \in \Delta$ is the unique data written, $\text{exp}$ is a numerical expression, and $D \in \mathcal{D}$ is the computation domain.
\end{mydef}

@@

In our formalism, it is forbidden for a stencil kernel to find $(r,n) \in R$ for which $r=w$ and $n \neq identity$. Actually, if such a property is possible implicit numerical schemes are needed into the simulation. Such a scheme is not a stencil and is over the scope of this paper. The problem does not happened for local kernels as $\forall (r,n) \in R$, $n = identity$. Computations for which this property is authorized are boundary kernels only. This type of kernel is a particular case as it commputes boundary conditions, which represents what should happened outside the space domain but still impact stencil kernels at the next time step.

@@

\medskip
A reduction computation is a computation which takes a single data applied onto a mesh and returns from all its entities a single data with a dimension reduced to zero, in other words a scalar. A reduction is typically used to compute the convergence criteria of the time loop of the simulation. Occasionally reductions can also be performed during a time iteration. %This last usage is typically done to have a conditionnal branch to choose one computation or another, or to compute a dynamic time step for example.

@@

\medskip
\noindent \textbf{Definitions}
\begin{itemize}
\item The set of $n$ ordered computation kernels of a numerical simulation is denoted $\Gamma = [k_i]_{0 \leq i \leq n-1}$, such that $\forall k_i,k_j \in \Gamma$, if $i \leq j$, then $k_i$ is computed before $k_j$.
\item Finally, a \textit{multi-stencil program} is defined by the octuplet 
\begin{equation*}
\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\mathcal{N},\Delta, \mathcal{S},\Gamma)
\end{equation*}
\end{itemize}

For example, in Figure~\ref{fig:ex1}, assuming that the computation domain (full lines) is denoted $dc1$ and the stencil shape is $n1$, the stencil kernel can be defined as:
\begin{equation*}
R: \{(B,n1)\}, \quad w: A, \quad d: dc1,
\end{equation*}
\begin{equation*}
exp: A(x,y)=B(x+1,y)+B(x-1,y)+B(x,y+1)+B(x,y-1).
\end{equation*}
On the other hand, in the example of Figure~\ref{fig:ex2}, assuming the computation domain is $dc2$ and the stencil shape is $n2$, the stencil kernel is defined as:
\begin{equation*}
R: \{(C,n2),(A,identity)\}, \quad w: A, \quad d: dc2,
\end{equation*}
\begin{equation*}
exp: A(x,y)=A(x,y)+C(x1,y1)+C(x1+1,y1).
\end{equation*}

@@

>>>>

We denote stencil kernels discretized explicit numerical schemes which involve neighborhood computations, whatever the type of space discrtization is (mesh).

@@

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item An entity $e$ of a mesh $\mathcal{M}=(V,E)$ is defined as a subset of its vertices and edges, $e\subset V\cup E$.
\item A group of mesh entities is denoted $E$ such that $E \in \mathcal{P}(V\cup E)$. It represents a set of entities of the same type.
\item The set of mesh entities groups of a simulation is denoted $\mathcal{E}$.
\item Finally, the mesh on which is applied a group of mesh entities $E$ is denoted $mesh(E)$
\end{itemize}
%\end{mydef}

For example, in a 2D Cartesian mesh an entity could be a cell, made of four vertices and four edges, or simply a vertex. As a result, a group of cells and a group of vertices could be defined in $\mathcal{E}$. %This type can be defined as the sets containing exactly four vertices and four edges connected as a cycle. Another type of entities simply are the vertices ($V$) and can be defined as all singletons formed of a single vertex of $V$.

@@

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item A quantity (or data) is a function $\delta: E_{\delta} \mapsto V_{\delta}$ which associates each entity of a group $e\in E_{\delta}$ to a value $v\in V_{\delta}$ where $V_{\delta}$ is a primitive data type, typically $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.
\item The set of quantities applied onto the mesh is denoted $\Delta$.
\item In the rest of this paper, the group of mesh entities on which a quantity $\delta$ is mapped is denoted $entity(\delta)=E_{\delta}$.
\end{itemize}
%\end{mydef}

Another option, closer to the applied mathematics domain, would have been to define $\delta$ as the function over $E_{\delta} \times T$. In this case a single quantity would represents all its occurences over time steps. This solution could be investigated in future work, however the approach chosen for now is to let the user be aware of the number of data he is using and what exactly for.

@@

In this section are considered four different types of kernel computations, stencil kernels, boundary kernels, local kernels, and reduction kernels.

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item A computation domain $D$ is a subpart of a mesh entities group, $D \subseteq E \in \mathcal{E}$.
\item The set of computation domains of a numerical simulation is denoted $\mathcal{D}$.
\item A neighborhood $n$ is a function which for a given entity $e \in E_i$, returns a set of $m$ entities in $E_j$, $n : E_i \rightarrow E_j^m$. One can notice that $i = j$ is possible. Most of the time, such a neighborhood is called a stencil shape.
\item The set of neihborhood functions in a numerical simulation is denoted $\mathcal{N}$.
\end{itemize}

\begin{mydef}
A computation kernel $k$ of a numerical simulation is defined as $k(S,R,(w,D),comp)$, where 
\begin{itemize}
\item $S \in \mathcal{S}$ is the set of scalar to read, 
\item $(w,D) \in \Delta \times \mathcal{D}$ is the single data modified by the computation kernel onto a given computation domain. Thus, $w \in \Delta$ and $D \in \mathcal{D}$ is the computation domain on which $w$ is computed, $D \subseteq E_w=entity(w)$.
\item $R$ is the set of tuples $(r,n)$, where $r \in \Delta$ and $n \in \mathcal{N}$ is a neighborhood function such that $n : E_w \rightarrow entity(r)^m$. The neighborhood indicates which entities of $r$ will be read in order to compute a single entity of $w$. 
\item $comp$ is the numerical computation which returns a value from a set of $n$ input values, $comp: V^n \rightarrow V$.
\end{itemize}
% \item A numerical expression $\text{exp}_{w}: entity(w) \times R^n \rightarrow V_{w}$ is a function which computes the values of the written quantity $w \in \Delta$ for a given entity $e \in entity(w)=E_w$, using a set of input data $R$.
% \item A computation kernel $k$ of a numerical simulation is defined as $c(R,w,\text{exp},D)$, where $R$ is the set of data read, $w \in \Delta$ is the unique data written, $\text{exp}$ is a numerical expression, and $D \in \mathcal{D}$ is the computation domain.
% \end{itemize}
%$R$ is the set of data read, $w \in \Delta$ is the unique data written, $\text{exp}$ is a numerical expression, and $D \in \mathcal{D}$ is the computation domain.
\end{mydef}

@@

In our formalism, it is forbidden for a stencil kernel to find $(r,n) \in R$ for which $r=w$ and $n \neq identity$. Actually, if such a property was permitted, implicit numerical schemes would be needed into the simulation, which involves linear solvers. Such a scheme is not a stencil and is over the scope of this paper. The problem does not happened for local kernels as $\forall (r,n) \in R$, $n = identity$. Computations for which this property is authorized are boundary kernels only. This type of kernel is a particular case as it commputes boundary conditions, which represents what should happened outside the space domain but still impact stencil kernels at the next time step.

@@

\medskip
A reduction computation is a computation which reads a single data applied onto a mesh and returns from all its entities a single scalar (mesh with a dimension reduced to zero). A reduction is typically used to compute the convergence criteria of the time loop of the simulation. Occasionally reductions can also be used during a time iteration. %This last usage is typically done to have a conditionnal branch to choose one computation or another, or to compute a dynamic time step for example.

@@

\medskip
\noindent \textbf{Definitions}
\begin{itemize}
\item The set of $n$ ordered computation kernels of a numerical simulation is denoted $\Gamma = [k_i]_{0 \leq i \leq n-1}$, such that $\forall k_i,k_j \in \Gamma$, if $i < j$, then $k_i$ is computed before $k_j$.
\item Finally, a \textit{multi-stencil program} is defined by the octuplet 
\begin{equation*}
\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\mathcal{N},\Delta, \mathcal{S},\Gamma)
\end{equation*}
\end{itemize}

For example, in Figure~\ref{fig:ex1}, assuming that the computation domain (full lines) is denoted $dc1$ and the stencil shape is $n1$, the stencil kernel can be defined as:
\begin{equation*}
R: \{(B,n1)\}, \quad w: A, \quad D: dc1,
\end{equation*}
\begin{equation*}
comp: A(x,y)=B(x+1,y)+B(x-1,y)+B(x,y+1)+B(x,y-1).
\end{equation*}
On the other hand, in the example of Figure~\ref{fig:ex2}, assuming the computation domain is $dc2$ and the stencil shape is $n2$, the stencil kernel is defined as:
\begin{equation*}
R: \{(C,n2),(A,identity)\}, \quad w: A, \quad D: dc2,
\end{equation*}
\begin{equation*}
comp: A(x,y)=A(x,y)+C(x1,y1)+C(x1+1,y1).
\end{equation*}

@@

One can note that all definitions given in this section are not dependent from the topology of the mesh. This property will be kept in the rest of this paper to propose the mesh-agnostic MSL language.

### 2016-02-01 ARTICLES/IJPP/article/msl.tex
article IJPP corrections+example Helene 01/02/2016

\begin{filecontents*}{exmsl.txt}
mesh : cart
mesh entities : cell, edgex
computation domains :
  d1 in cell
  d2 in edgex
independent :
  d1 and d2
stencil shapes : 
  n1 from cell to cell
  n2 from cell to edgex
  n3 from edgex to cell
data :
  A,cell
  B,cell
  C,edgex
  D,cell
  E,cell
  F,cell
  G,cell
  H,edgex
  I,cell
  J,cell
scalar : mu, tau
time : 500
computations :
  B[d1] = k0({tau},{A})
  C[d2] = k1({},{B[n2]})
  D[d1] = k2({},{C})
  E[d1] = k3({},{C})
  F[d1] = k4({},{D,C[n3]})
  G[d1] = k0({mu,tau},{E})
  H[d2] = k6({},{F})
  I[d1] = k7({},{G,H})
  J[d1] = k8({mu},{I[n1]})
\end{filecontents*}

>>>>

\begin{filecontents*}{exmsl.txt}
mesh : cart
mesh entities : cell, edgex
computation domains :
  d1 in cell
  d2 in edgex
independent :
  d1 and d2
stencil shapes : 
  ncc from cell to cell
  nce from cell to edgex
  nec from edgex to cell
data :
  A,cell
  B,cell
  C,edgex
  D,cell
  E,cell
  F,cell
  G,cell
  H,edgex
  I,cell
  J,cell
scalar : mu, tau
time : 500
computations :
  B[d1] = k0({tau},{A})
  C[d2] = k1({},{B[n2]})
  D[d1] = k2({},{C})
  E[d1] = k3({},{C})
  F[d1] = k4({},{D,C[n3]})
  G[d1] = k0({mu,tau},{E})
  H[d2] = k6({},{F})
  I[d1] = k7({},{G,H})
  J[d1] = k8({mu},{I[n1]})
\end{filecontents*}

### 2016-02-01 ARTICLES/IJPP/article/msp.tex
article IJPP corrections Helene 29/01/2016
article IJPP corrections+example Helene 01/02/2016

%--------------------
\subsection{Minimal series-parallel DAG}
\begin{mydef}
A vertex $v$ of a DAG $G$ is a \emph{source} if no edge of $G$ enters $v$. Similarly, a vertex $v$ is a \emph{sink} if no edge of $G$ leaves $v$. 
\end{mydef}

In 1982, Valdes \& Al~\cite{Valdes:1979:RSP:800135.804393} have defined the class of minimal series-parallel DAGs (MSPD).

@@

\begin{proof}
Actually, it is possible in a multi-stencil program to have a set of computations such that their dependencies form the forbidden $N$ subgraph. Considering four computation kernels $k_0$, $k_1$, $k_2$ and $k_3$ of a $\mathcal{MSP}$ such that $k_0 \prec k_1$, $k_2 \prec k_1$ and $k_2 \prec k_3$ are verified. Thus, the \emph{zigzag} relation $k_0 \prec k_1 \succ k_2 \prec k_3$ which form a forbidden subgraph of a MSPD is found in $\Gamma_{dep}$.
\end{proof}

@@

The algorithm is inspired from the one proposed by~\cite{Mitchell:2004:CMV:1082101.1082117}, and is described in Algorithm~\ref{alg:nshape}. We precisely describe how we find edges to add to get rid of N-Shapes. One can notice that another solution would have been to remove the extra-edge from the N-Shape instead of adding one~\cite{}. However, we have chosen to exactly keep dependencies computed in $\Gamma_{dep}$, to not approximate the static scheduling. Finally, algorithm proposed by~\cite{Bender200575} to find a Lowest Common Ancestor in a DAG (in $\mathcal{O}(|V|^3)$, could have been also used.

The first step of Algorithm~\ref{alg:nshape} is to create a new DAG $\Gamma_{dep}'$ from $\Gamma_{dep}$, such that vertices having exactly one source and one destination in $\Gamma_{dep}$ is not added to $\Gamma_{dep}'$, and its source and destination vertices are directly linked by an edge. This step is called $RTV(\Gamma_{dep})$ (line 2) and is done in $\mathcal{O}(|V|)$. The second step is to get root vertices in $\Gamma_{dep}'$. The third step calls $BFSnshape$ (line 4). This procedure is based on a Breadth First Search recursive algorithm (in $\mathcal{O}(|V|+|E|)$) and takes as input a list of source vertices $sources$, $\Gamma_{dep}$ and $\Gamma_{dep}'$. At each recursive call the induced sub-graph of $\Gamma_{dep}'$ containing $sources$ vertices and the list of their successors $successors(sources)$ is created and called $SG$. If this sub-graph $SG$ is connected and is not a complete bipartite graph, the algorithm make the subgraph $SG$ complete by adding needed edges into $\Gamma_{dep}'$ (line 12 in $\mathcal{O}(|V|^2$). The algorithm also add same edges to $\Gamma_{dep}$ (line 13). Finally, line 5, the transitive closure of $\Gamma_{dep}$ is computed.

\begin{algorithm}
\caption{Remove N-Shapes}
\label{alg:nshape}
\begin{algorithmic}[1]
\Procedure{removeNSH} {$\Gamma_{dep}$}
\\$\Gamma_{dep}'$ = RTV($\Gamma_{dep}$)
\\$sources$ = roots($\Gamma_{dep}'$)
\\BFSnshape($sources$,$\Gamma_{dep}'$,$\Gamma_{dep}$)
\\TC($\Gamma_{dep}$)
\EndProcedure

\Procedure{BFSnshape} {$sources$, $\Gamma_{dep}'$, $\Gamma_{dep}$}
\\$successors$ = successors($sources$)
\\$SG$ = subgraph($sources \cup successors$)
\If{isconnected($SG$)}
\If{! complete($SG$)}
\\addedges($\Gamma_{dep}'$)
\\addedges($\Gamma_{dep}$)
\\BFSnshape($successors$, $\Gamma_{dep}'$, $\Gamma_{dep}$)
\EndIf
\If{! isconnected($SG$)}
\\BFSnshape($successors$, $\Gamma_{dep}'$, $\Gamma_{dep}$)
\EndIf
\EndIf
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

>>>>

%--------------------
\subsection{Minimal Series-Parallel DAG}
\begin{mydef}
A vertex $v$ of a DAG $G$ is a \emph{source} if no edge of $G$ enters $v$. Similarly, a vertex $v$ is a \emph{sink} if no edge of $G$ leaves $v$. 
\end{mydef}

In 1982, Valdes \& Al~\cite{Valdes:1979:RSP:800135.804393} have defined the class of Minimal Series-Parallel DAGs (MSPD).

@@

\begin{proof}
Actually, it is possible in a multi-stencil program to have a set of computations such that their dependencies form the forbidden $N$ subgraph. For example, considering four computation kernels $k_0$, $k_1$, $k_2$ and $k_3$ of a $\mathcal{MSP}$ such that $k_0 \prec k_1$, $k_2 \prec k_1$ and $k_2 \prec k_3$ are verified. The \emph{zigzag} relation $k_0 \prec k_1 \succ k_2 \prec k_3$ which form a forbidden subgraph of a MSPD is found in $\Gamma_{dep}$.
\end{proof}

@@

The algorithm is inspired from the one proposed by~\cite{Mitchell:2004:CMV:1082101.1082117}, and is described in Algorithm~\ref{alg:nshape}. We precisely describe how we find edges to add to get rid of N-Shapes. One can notice that another solution would have been to remove the extra-edge from the N-Shape instead of adding one~\cite{}. However, we have chosen to exactly keep dependencies computed in $\Gamma_{dep}$, to not approximate the static scheduling. Finally, algorithm proposed by~\cite{Bender200575} to find a Lowest Common Ancestor in a DAG (in $\mathcal{O}(|V|^3)$, could have also been used.

The first step of Algorithm~\ref{alg:nshape} is to create a new DAG $\Gamma_{dep}'$ from $\Gamma_{dep}$, such that vertices having exactly one source and one destination in $\Gamma_{dep}$ are not added to $\Gamma_{dep}'$, and its source and destination vertices are directly linked by an edge. This step is called $RTV(\Gamma_{dep})$ (line 2) and is done in $\mathcal{O}(|V|)$. The second step is to get root vertices in $\Gamma_{dep}'$. The third step calls $BFSnshape$ (line 4). This procedure is based on a Breadth First Search recursive algorithm (in $\mathcal{O}(|V|+|E|)$) and takes as input a list of source vertices $sources$, $\Gamma_{dep}$ and $\Gamma_{dep}'$. At each recursive call the induced sub-graph of $\Gamma_{dep}'$ containing $sources$ vertices and the list of their successors $successors(sources)$ is created and called $SG$. If this sub-graph $SG$ is connected and is not a complete bipartite graph, the algorithm make the subgraph $SG$ complete by adding needed edges into $\Gamma_{dep}'$ (line 12 in $\mathcal{O}(|V|^2$). The algorithm also add same edges to $\Gamma_{dep}$ (line 13). Finally, line 5, the transitive closure of $\Gamma_{dep}$ is computed.

\begin{algorithm}
\caption{Remove N-Shapes}
\label{alg:nshape}
\begin{algorithmic}[1]
\Procedure{removeNSH} {$\Gamma_{dep}$}
\State $\Gamma_{dep}'$ = RTV($\Gamma_{dep}$)
\State $sources$ = roots($\Gamma_{dep}'$)
\State BFSnshape($sources$,$\Gamma_{dep}'$,$\Gamma_{dep}$)
\State TC($\Gamma_{dep}$)
\EndProcedure

\Procedure{BFSnshape} {$sources$, $\Gamma_{dep}'$, $\Gamma_{dep}$}
\State $successors$ = successors($sources$)
\State $SG$ = $\Gamma_{dep}'$.subgraph($sources \cup successors$)
\If{isconnected($SG$)}
\If{! complete($SG$)}
\State addedges($\Gamma_{dep}'$)
\State addedges($\Gamma_{dep}$)
\State BFSnshape($successors$, $\Gamma_{dep}'$, $\Gamma_{dep}$)
\ElsIf{! isconnected($SG$)}
\State BFSnshape($successors$, $\Gamma_{dep}'$, $\Gamma_{dep}$)
\EndIf
\EndIf
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

Proof of the algorithm and complexity improvments could be studied in future works. However, as it will be shown in Section~\ref{sect:eval}, the complexity of this algorithm is not a problem because of the relatively small size of $\Gamma_{dep}$.
